Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-08-31 bbbq_article.tex
Show percentage TMH expected, fix #39
Remove tedious evolution experiment

\begin{itemize}
  \item $\mathcal{H}_{1,VB}$: MHC-I has the same percentage of epitopes overlapping
    with viral/bacterial TMHs as with human TMHs
  \item $\mathcal{H}_{2,HVB}$: MHC-II has the same percentage of epitopes overlapping
    with human/viral/bacterial TMHs as expected by chance
  \item $\mathcal{H}_{3}$: The mutations observed in 
    viral TMHs are caused by chance.
    Alternatively, these are caused by natural selection
    induced by (non-)binding to HLAs
\end{itemize}

@@

To determine the percentages of epitopes overlapping
with TMHs, we use three reference proteomes, which are
a human, viral and bacterial one.
The human reference proteome is \richel{here}, 
which is the same as \cite{bianchi2017}.
The viral reference proteome is represented by
the proteome of the first sequenced COVID-19 strain (\cite{wu2020new},
GenBank ID of MN908947.3, \url{https://www.ncbi.nlm.nih.gov/nuccore/MN908947})
For bacteria, the reference genome of 
Mycobacterium tuberculosis (\url{https://www.ebi.ac.uk/reference_proteomes}, UP000001584, 
83332 MYCTU) is used.

\paragraph{Data set for evolutionary selection}

To detect an evolutionary signal for selection on binding
TMHs peptides, we need a phylogenetic tree with 
many time-dated proteomes.
To do so, we use a viral pathogen, as these are detected by both
MHC-I and MHC-II.
To detect individual mutations, we need a high resolution dataset
that also includes the evolutionary history of the virus.
COVID-19 satisfies both requirements:
as of 2020-06-22, it has 49790 proteomes, as provided by GISAID (\cite{shu2017gisaid}),
as well as a provided evolutionary history by Nextstrain (\cite{hadfield2018nextstrain}).

@@

This study differs in some aspects, described below in more detail,
which is the definition of what a binder is,
the addition of using MHC-II haplotypes, the software used,
and the significance level to determine if TMH-derived peptides are as 
likely to be a binder as expected by chance.
These deviations are either a refinement of the previous method or
a pragmatic choice made due to the extension of the original experiment.

The definition of a binder differs from \cite{bianchi2017}:
in this study a peptide is labeled an MHC-I binder if, within a haplotype, 
any of its 9-mer peptides have an IC50 in the lowest 2\% of randomly simulated 
peptides (see table \ref{table:ic50s_per_haplotype} for values), were the original study defined
a binder as having an IC50 in the lowest 2\% of the peptides within a protein.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17

The 13 MHC-I haplotypes used in this study are the same as 
the ones in \cite{bianchi2017}.
The MHC-II haplotypes used additionally are selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in 13 haplotypes.
This frequency threshold is arbitrarily chosen to result in 
the same number of haplotypes for each MHC class
\richel{Add this to bbbq, there it is now 21}.
When using an MHC-II haplotype, instead of using 9-mers, 13-mers are
used, as these are the canonical MHC-II epitope size \richel{Reference here}.

Instead of using the MHC-I-only epitope prediction 
tool \verb;EpitopePrediction;, MHCnuggets is used, as it can predict
both MHC-I and MHC-II epitopes.
Additionally, instead of only using a human proteome, in this study
also a viral and bacterial proteome are analyzed.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Detecting evolution}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To detect evolution, we determine the probability 
that all of the observed TMH mutations have arisen by chance,
with regards to being detected/presented by the immune system,
per MHC class and haplotype. We do this for human, viral and
bacterial TMHs.

\paragraph{Correction against many tests}

% Chance to find
%            this amount of positives (sums to 1)
%              |
%              v
% > dbinom(x = 0, size = 26, prob = 0.05)
% [1] 0.2635201
% > dbinom(x = 1, size = 26, prob = 0.05)
% [1] 0.3606064
% > dbinom(x = 2, size = 26, prob = 0.05)
% [1] 0.2372411
% > dbinom(x = 3, size = 26, prob = 0.05)
% [1] 0.09989098
% > dbinom(x = 4, size = 26, prob = 0.05)
% [1] 0.03023017
% > dbinom(x = 5, size = 26, prob = 0.05)
% [1] 0.00700067
%
% Chance to find more than
%            this amount of positives (excluding the value!)
%              |
%              v
% > pbinom(q = 0, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.7364799
% > pbinom(q = 1, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.3758735
% > pbinom(q = 2, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.1386324
% > pbinom(q = 3, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.0387414
% > pbinom(q = 4, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.008511231

Because we use 13 haplotypes per MHC class,
for a significance level of 5\%,
for both directions (D to U, or U to D)
we expect 5% times 26 equals 1.3 significant findings.

\richel{Hmm, improve}
The chance to find 2 haplotypes that hint to have selection working upon it
is 4\%, which is below our significance level.
% > pbinom(q = 3, size = 26, prob = 0.05, lower.tail = FALSE)
% [1] 0.0387414

Upon finding 3 haplotypes that hint to have selection working upon it,
there is the additional chance that selection is working in different
directions (i.e. in one case there is selection for losing detection,
where in the other case, there is (unexpected) selection for gaining
detection). Therefore, finding 3 haplotypes that hint to have selection working 
upon it in the same direction

Therefore, we will claim there is directional selection, 
when, for a proteome, we find 2 haplotypes 
that hint to have selection working upon it

\paragraph{System depicted as a Markov chain}

To calculate the probability all of our observations happened by chance,
we view our experiment as a two-state Markov chain, as depicted in
figure \ref{fig:markov_chain}. The two nodes depict the
two possible detection state a peptide is in, which is
either undetected $U$, or detected $D$ by the immune system,
where a detected peptide is a binder, that is, 
a peptide with an IC50 in the lowest 2\%.
The $D$ can be either detection by an MHC-I or MHC-II haplotype.
The edges depict the probability of a state transition by mutation,
for example, $p_{DU}$ denotes the probability that a detected
peptide mutates into an undetected one.
We focus on peptides that are TMH and we assume a mutation that 
renders the TMH non-functional as lethal.

\begin{figure}
  \centering
  \input{markov_chain_2_states.latex}
  \caption{
    State transitions in TMHs caused by mutations, with respect to the
    detection by the immune system. Any TMH is either in a detected, $D$, or
    undetected, $U$, state. Mutations cause state transitions with probability
    $p$. For example, $p_{DU}$ denotes the probability that a detected
    TMH mutates into an undetected one.
  }
  \label{fig:markov_chain}
\end{figure}

@@

When we have calculated the probability of each transition (as described below),
we can simply count the number of observed transitions.
Because the probability of transitions differ per peptide,
the total system has the properties of a Poisson binomial distribution,
from which we can calculate the probability of the observing our
number of transitions.

\paragraph{Hypotheses testing}

Of the complete system, we test for each of the 2 observed 
state transition rates (that is, we do not test for mutations that result in the same state)
if these are likely to be caused by chance.
For each transition rate, we know the expected and observed number
of transitions. If the observed number of transitions is below the expected
value, we test for finding that (low) number of transitions or less.
If the observed number of transitions is above the expected
value, we test for finding that (high) number of transitions or more.
We use a significance level $\alpha$ of 5\%, because we have no prior
expectations.

As we are testing 2 rates (and thus 2 hypotheses), 
we correct against type I errors (false positives), 
by applying the Bonferroni correction, 
resulting in $\alpha_c = \frac{\alpha}{2}$.
If the probability of observing that few/many 
transition is below $\alpha_c$, we will state 
that this unlikely situation may be caused by natural selection.

\paragraph{Calculating one peptide's probability}

We calculate the probability of a transition per peptide.
For any peptide, we can generate all sequences that differ
in only one amino acids. Of these sequences, we only keep the ones
that result in a TMH. Of each of the remaining TMH sequences, we predict
which mutants are detected by the immune system. Because not all 
AA mutations are equally likely, we correct for the AA transition
rates using the FLU transition matrix (\cite{dang2010flu}),
which is a transition matrix derived from observed influenza AA mutations.

\paragraph{Assumption that epitopes are immunogenic}

We assume that if a peptide binds to an MHC, it is always presented and always
leads to an immune response.
The more often this assumption is violated, 
the more our analysis will be weakened.
Experiments in mice with MHC-I epitopes, however, indicate that (if the dose 
administered was high enough) there are either none or rare
exceptions \cite{sette1994relationship}.

\paragraph{Measuring strength}

To assess the strength of our analysis, 
we have applied it to four simulated data sets,
which are the product of using either MHC-I or MHC-II,
as well as using two different selection scenarios.

For both MHC-I and MHC-II, we use the haplotypes that have 
the lowest phenotypic frequency, which is \richel{X} for MHC-I
and DPB1*0501 (21.7\%) for MHC-II.

The two selection scenarios are either no selection
or a two-fold selection on what we would expect.
The scenario without selection, the 'null' data set, acts 
as a control, where we let the simulate observations occur by chance. 
The 'rigged' data set is a data set in which we 
let evolution bias the observations two-fold, that is, 
we let mutations that are beneficial occur twice as often as expected by chance.

All data sets consists out of $10^4$ observations.
Each observation starts with a randomly generated source peptide fragment,
as well a mutant. For the 'null' data set, the mutant is randomly select.
For the 'rigged' data set, beneficial mutations are twice as likely to
be selected. We define a beneficial mutation as a transition from
being detected to being undetected.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Prediction software used}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

There are multiple computational tools developed to predict which
parts of membrane proteins are TMH
In 2001, multiple tools to do so have been compared \cite{moller2001evaluation},
of which TMHMM \cite{krogh2001predicting} turned out to be the best, yet
this software has a restrictive software license and is nearly two
decades old. Inspired on TMHMM, many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}).
The most recent FOSS tool is called PureseqTM (\cite{wang2019efficient}),
which is used in this study.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons \cite{karosiene2012netmhccons} gave the best predictions.
A tool developed later is \verb;epitope-prediction; \cite{bianchi2017},
which uses a stabilized matrix method \cite{kim2009derivation}.
The most recent FOSS tool available is MHCnuggets\cite{shao2020high},
which we will use in this study

@@

% \paragraph{Gene prediction}
% 
% Proteogenomics is the field dedicated to, among others,
% predict the proteins that arise from a DNA 
% sequence \cite{nesvizhskii2014proteogenomics}.
% For an ab initio gene prediction of a viral genome, 
% we searched the literature for the most recent software tool,
% which is Vgas (\cite{zhang2019vgas}). In this article, 
% Vgas is compared to tools, such as GeneMarkS (\cite{besemer2001genemarks}),
% Prodigal (\cite{hyatt2012gene}), GLIMMER (\cite{delcher1999improved}),
% RAST \cite{aziz2008rast} and prokka (\cite{seemann2014prokka}).
% We picked the FOSS tool that scored best in the (closed source) Vgas
% benchmark (\cite{zhang2019vgas}), which is Prodigal.

@@

The R programming language is used for the complete 
experiment, including the analysis.
The complete experiment is bundled in the 'bbbq' R package,
which is dependent on 'pureseqtmr', 'mhcnuggetsr' and 'mhcnpreds',
as described below.

@@

To be able to use pre-calculated MHCnuggets
predictions (i.e. a distribution of IC50 values per haplotype, 
derived from random peptides), the 'mhcnpreds' R package has been 
developed. In practice, this is nothing more than a big look-up table.
Because the package is too big, it will not been submitted to CRAN.

@@

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{ic50s_per_haplotype/ic50s_per_haplotype.png}
  Predicted IC50s (in nM) per haplotype
  \caption{
    \richel{Used randomly simulated peptides, results are real}
  }
  \label{fig:ic50s_per_haplotype}
\end{figure}

>>>>

\begin{itemize}
  \item $\mathcal{H}_{1,VB}$: MHC-I has the same percentage of epitopes overlapping
    with viral/bacterial TMHs as with human TMHs
  \item $\mathcal{H}_{2,HVB}$: MHC-II has the same percentage of epitopes overlapping
    with human/viral/bacterial TMHs as expected by chance
  \item $\mathcal{H}_{3}$: TMH and non-TMH regions are equally conserved
    in a proteome
\end{itemize}

@@

To determine the percentages of epitopes overlapping
with TMHs, we use three reference proteomes, which are
a human, viral and bacterial one.
For a human reference proteome, we used the Uniprot human reference proteome
from \url{https://www.uniprot.org/proteomes/UP000005640}), 
which is the same as \cite{bianchi2017}.
The viral reference proteome is represented by
the proteome of the first sequenced COVID-19 strain (\cite{wu2020new},
GenBank ID of MN908947.3, \url{https://www.ncbi.nlm.nih.gov/nuccore/MN908947})
For bacteria, we used the Uniprot Mycobacterium tuberculosis reference genome
(\url{https://www.uniprot.org/proteomes/UP000001584}) is used.

\paragraph{Data set for evolutionary conservation of TMHs}

To detect the evolutionary conservation of TMHs,
we use the 49790 (as of 2020-06-22) fully sequenced SARS-CoV-2 proteomes, 
as provided by GISAID (\cite{shu2017gisaid}).

@@

This study differs in some aspects, described below in more detail,
which is the definition of what a binder is,
the addition of using MHC-II haplotypes, the software used,
and the significance level to determine if TMH-derived peptides are as 
likely to be a binder as expected by chance.
These deviations are either a refinement of the previous method or
a pragmatic choice made due to the extension of the original experiment.
Additionally, instead of only using a human proteome, in this study
also a viral and bacterial proteome are analyzed.

The definition of a binder differs from \cite{bianchi2017}:
in this study a peptide is called a binder if, within a haplotype, 
any of its 9-mer peptides have an IC50 value in the lowest 2\% of 
one thousand randomly simulated 
peptides (see table \ref{table:ic50s_per_haplotype} for values), 
whereas the original study defined
a binder as having an IC50 in the lowest 2\% of the peptides within a protein.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17

The 13 MHC-I haplotypes used in this study are the same as 
the ones in \cite{bianchi2017}.
The MHC-II haplotypes used additionally are selected 
to occur with a phenotypic frequency of at least 
\richel{recalculate number}
14\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in 21 haplotypes.
When using an MHC-II haplotype, instead of using 9-mers, 13-mers are
used, as these are the canonical MHC-II epitope size \richel{Reference here}.

@@

Instead of using the membrane protein topology prediction
tool \verb;TMHMM; \cite{krogh2001predicting}, PureseqTM \cite{wang2019efficient}
is used instead.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Evolutionary conservation of TMHs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

To detect the evolutionary conservation of TMHs,
we create an alignment using msa, that uses msaClustalW in the back,
create a consensus sequence, determine the topology of the consensus
using the FLU transition matrix,
measure the topology of it, then estimate the conservation values.

@@

We compare the conservation values for TMH versus non-TMH locations,
using a Kolmogorov-Smirnov test, for a significance level of 5 percent.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Prediction software used}
\label{subsec:prediction_software_used}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{table}[]
  \begin{tabular}{lll}
    Goal & Tool  \\ 
    \hline
    Predict topology                  & TMHMM \cite{krogh2001predicting} \\
    Predict topology                  & PureseqTM \cite{wang2019efficient} \\
    Predict epitopes MHC-I            & \verb;epitope-prediction; \cite{bianchi2017} \\
    Predict epitopes MHC-II           & NetMHCIIpan \cite{nielsen2008quantitative,karosiene2013netmhciipan} \\
    Predict epitopes MHC-I and MHC-II & MHCnuggets \cite{shao2020high} \\
    Call TMHMM from R                 & tmhmm \\
    Call PureseqTM from R            & pureseqtmr \\
    Call NetMHCIIpan from R           & netmhc2pan \\
    Call MHCnuggets from R            & mhcnuggetsr \\
    Call TMHMM from R                 & tmhmm \\
    \verb;epitope-prediction; LUT     & epiprepreds \\
    NetMHCIIpan LUT                   & nmhc2ppreds \\
    MHCnuggets LUT                    & mhcnpreds \\
    Combine all                       & bbbq
  \end{tabular}
  \caption{
    Overview of all software used in this research.
    Software without a reference is written for this research.
    'LUT' is an abbreviation for 'Look-up table'. 
  }
  \label{table:software_used}
\end{table}

@@

There are multiple computational tools developed to predict which
parts of membrane proteins are TMH.
In 2001, multiple tools to do so have been compared \cite{moller2001evaluation},
of which TMHMM \cite{krogh2001predicting} turned out to be the best.
TMHMM is used in the previous study \cite{bianchi2017},
yet TMHMM itself has a restrictive software license and is nearly two
decades old. 
Inspired on TMHMM, many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}).
The most recent FOSS tool is called PureseqTM (\cite{wang2019efficient}),
which is used in this study.

@@

For MHC-I, there are multiple computational tools developed 
to predict epitopes. 
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCcons \cite{karosiene2012netmhccons} gave the best predictions.
A tool developed later is \verb;epitope-prediction; \cite{bianchi2017},
which uses a stabilized matrix method \cite{kim2009derivation}.
The most recent FOSS tool available is 
MHCnuggets \cite{shao2020high} (which can do both MHC-I 
and MHC-II prediction) which we will use in this study

@@

@@

The R programming language is used for the complete 
experiment, including the analysis.
The complete experiment is bundled in the 'bbbq' R package,
which is dependent on 'pureseqtmr', 
'mhcnuggetsr' and 'mhcnpreds',
'epitope-prediction' and 'epiprepreds',
'netmhc2pan' and 'nmhc2ppreds'
as described below.

@@

To be able to use pre-calculated 
\verb;epitope-prediction;, NetMHCIIpan and MHCnuggets
predictions (i.e. a distribution of IC50 values per haplotype, 
derived from random peptides), the 'epiprepreds', 'nmhc2ppreds' 
and 'mhcnpreds' R package have been 
developed. In practice, these are nothing more than a big look-up table.
If \verb;epitope-prediction; is put onto CRAN, 'epiprepreds'
will be submitted as well.
If 'netmhc2pan' is allowed onto CRAN, 'nmhc2ppreds'
will be submitted as well.
If 'mhcnuggetsr' is put onto CRAN, 'mhcnpreds'
will be submitted as well.

@@

\input{bbbq_1/table_f_tmh.latex}

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.9\textwidth]{ic50s_per_haplotype/ic50s_per_haplotype.png}
  Distribution of predicted IC50s (in nM) per haplotype
  \label{fig:ic50s_per_haplotype}
\end{figure}
